---
title: "Chat GPT"
author: "Erik Westlund"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(causaldata)
library(ggrepel)
library(showtext)

source("colors.R")

source_auto <- auto
auto <- source_auto |> 
  mutate(
    weight_cat = cut_number(weight, 2, labels = c("Compact", "Full")),
    manufacturer =   sapply(str_split(make, " "), `[`, 1)
  )

```

# Chat GPT

**Prompt:**

I have a plot of auto data. The data set is named "auto." The variables I'm interested are: price, efficiency (variable name: mpg), weight, and manufacturer.  The data set only includes the variable "make," which has both the manufacturer and the car name.  Can you first split "make" into "manufacturer" and "model." Weight is a continuous measure, but it would be helpful to split this data into categories of small, medium, and large.

I'd then like a plot comparing efficiency to weight. Facet this into small, medium, and large cars. On each figure, annotate the least efficient car with its make and model, one average car with its make and model, and the most efficient car with its make and model.

Use ggplot2 with the minimal theme. Change the type face to "Inter"

**Result:**

```{r chat_gpt}
# Load the Inter font
font_add_google("Inter", "Inter")
showtext_auto()

# Prepare the data
auto <- source_auto |>
  # Split "make" into "manufacturer" and "model"
  separate(make, into = c("manufacturer", "model"), sep = " ", extra = "merge") |>
  # Categorize weight into small, medium, and large
  mutate(weight_category = cut(weight,
                               breaks = c(-Inf, 3000, 4000, Inf),
                               labels = c("Small", "Medium", "Large"))) |>
  # Group data for annotations
  group_by(weight_category) |>
  mutate(
    least_efficient = mpg == min(mpg),
    most_efficient = mpg == max(mpg),
    average_efficient = mpg == median(mpg)
  ) |>
  ungroup()

# Create the plot
ggplot(auto, aes(x = weight, y = mpg)) +
  geom_point(aes(color = manufacturer), alpha = 0.7) + # Points colored by manufacturer
  facet_wrap(~weight_category, scales = "free_x") + # Facet by weight category
  geom_text(
    data = auto |> filter(least_efficient),
    aes(label = paste(manufacturer, model)),
    hjust = 1.2, vjust = -0.5, color = "red"
  ) + # Annotate least efficient cars
  geom_text(
    data = auto |> filter(average_efficient),
    aes(label = paste(manufacturer, model)),
    hjust = 1.2, vjust = -0.5, color = "blue"
  ) + # Annotate average cars
  geom_text(
    data = auto |> filter(most_efficient),
    aes(label = paste(manufacturer, model)),
    hjust = 1.2, vjust = -0.5, color = "green"
  ) + # Annotate most efficient cars
  labs(
    title = "Efficiency vs Weight by Car Size",
    x = "Weight",
    y = "Miles Per Gallon (MPG)"
  ) +
  theme_minimal(base_family = "Inter") +
  theme(legend.position = "bottom")

ggsave("images/chat_gpt_1.png", width = 12, height = 5.626, dpi = 300)

```

**Follow-up:**

That's a good start, but there are issues.  For one, the text colors look bad. Can you pick out a color from Tailwind's design system, in the 600 or 700 range?  Moreover, the labels need to dodge the points. They are sometime son top of each other. Moreover, it looks like you added more than one car per facet. Please don't do this.  There is too much information about manufacturers.   So please remove per-manufacturer colors and that legend entirely.  Could you add a GAM smoother to each plot? Center the title and make it bold.  Make sure we only have three car labels per facet and make sure they are never on top of each other.

**I told it to fix a few errors and then told it:**

A few issues. There are not a lot of large cars.  Lets start over.  By the way, I stored the source auto_data as "source_auto" so you can always start any data manipulations with source_data.  

Let's keep this idea but fix the following things.  Lets do "compact cars" and "large cars". So only two facets.

Please, add only a single label for each of these categories:  Least, average, and most efficient.

Can you make the smoothing ribbon a lighter gray? It's distracting.

**And then:**

okay, good start.  Change "Average efficient" to "Average efficiency"

And stop having so many average groups.  There's correctly one least and one most.  But there are something like five of average efficiency.  I believe this is because there are multiples with the same value.  Please make sure it picks only one.

**And then...**

The text for average is hard to read because its on the line and ribbon. Can we dodge that?  We can put the labels farther away and point at them with arrows.

**And...**

There are ties for least efficient, so you are ending up with two there. You need to do the same thing for least, average, and most.   There are no arrows; only for one on least.   Can you also make those specific points green, blue, and red (most, average, least) the rest of the points can be a lighter gray.  

**And...*

Almost there. Make the points a little darker. You can use tailwind slate 500

then use green 600 for most and red 700 for least. 

Change the  "vs" in the title to "vs." Make the facet headings bold.  And make the x and y axis labels bold.

**And...**

Very close.  Make the average dot a blue 700 from tailwind.  It's hard to see.  Make the weight say Weight (LBS.) and make the smoother ribbon just a little less dark.

**And...**

OK. Very close.  Change "LBS" to "lbs."  Make the blue dot darker.  And change the title to "Efficiency vs. Weight for Compact and Large Cars"

**And..."

change "Large Cars" to "Full-size Cars"

```{r chat_gpt_2}
# Load the Inter font
font_add_google("Inter", "Inter")
showtext_auto()

# Start with the original dataset
auto <- source_auto |>
  # Split "make" into "manufacturer" and "model"
  separate(make, into = c("manufacturer", "model"), sep = " ", extra = "merge") |>
  # Create categories for compact and full-size cars
  mutate(weight_category = cut(weight,
                               breaks = c(-Inf, 3000, Inf),
                               labels = c("Compact Cars", "Full-size Cars")))

# Get least efficient cars per category (single tie-resolved row)
least_efficient <- auto |>
  group_by(weight_category) |>
  slice_min(mpg, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(label = paste(manufacturer, model, "(Least Efficient)"), color = "#B91C1C") # Red 700

# Get most efficient cars per category (single tie-resolved row)
most_efficient <- auto |>
  group_by(weight_category) |>
  slice_max(mpg, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(label = paste(manufacturer, model, "(Most Efficient)"), color = "#16A34A") # Green 600

# Get average efficiency cars per category (single tie-resolved row)
average_efficient <- auto |>
  group_by(weight_category) |>
  slice_min(abs(mpg - mean(mpg, na.rm = TRUE)), n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(label = paste(manufacturer, model, "(Average Efficiency)"), color = "#1E40AF") # Blue 800

# Combine all annotations
annotations <- bind_rows(least_efficient, most_efficient, average_efficient)

# Plot
ggplot(auto, aes(x = weight, y = mpg)) +
  # All other points in slate 500
  geom_point(color = "#64748B", alpha = 0.8) + # Slate 500
  # Highlighted points for least, average, and most efficient
  geom_point(
    data = annotations,
    aes(color = color), size = 3
  ) +
  scale_color_identity() + # Use colors directly from data
  # GAM smoother with lighter ribbon
  geom_smooth(
    method = "gam", formula = y ~ s(x, bs = "cs"),
    color = "#334155", fill = "gray90", alpha = 0.4
  ) +
  # Labels with arrows pointing to their points
  geom_label_repel(
    data = annotations,
    aes(label = label),
    fill = "white", # White background for contrast
    color = "#334155",
    family = "Inter",
    size = 4, # Increase label size
    box.padding = 0.5,
    point.padding = 0.5,
    arrow = arrow(length = unit(0.02, "npc")), # Arrows pointing to labels
    max.overlaps = Inf
  ) +
  # Two facets: Compact and Full-size Cars
  facet_wrap(~weight_category, scales = "free_x", labeller = label_value) +
  labs(
    title = "Efficiency vs. Weight for Compact and Full-size Cars",
    x = "Weight (lbs.)",
    y = "Miles Per Gallon (MPG)"
  ) +
  theme_minimal(base_family = "Inter", base_size = 16) + # Increase base text size
  theme(
    legend.position = "none", # Remove legend
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20), # Larger title
    strip.text = element_text(face = "bold", size = 18), # Larger facet headings
    axis.title = element_text(face = "bold", size = 16), # Larger axis labels
    axis.text = element_text(size = 14), # Larger axis text
    panel.grid.minor = element_blank()
  )

ggsave("images/chat_gpt_2.png", width = 12, height = 5.625, dpi=300)

```
